---
title: "Introduction to Stray::Pibble"
author: "Justin Silverman"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{intro_stray_pibble}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# An introduction to *stray*
*stray* [@silverman2019] is a loose acronym for **"(Bayesian) Multinomial Logistic-Normal Models"**. 
In particular the development of 
*stray* stems from the need for fast inference for time-invariant MALLARD models[@silverman2018]. 
*stray* is very fast! It uses closed form solutions for model gradients and Hessians written
in C++ to preform [MAP estimation](https://en.wikipedia.org/wiki/Maximum_a_posteriori_estimation)
in combination with parameter uncertainty estimation using a [Laplace Approximation](http://www.sumsar.net/blog/2013/11/easy-laplace-approximation/).
One of the main models in *stray* is the function *pibble* which fits a Multinomial
Logistic-Normal **Linear Regerssion** model. 

**So what is a *stray* model exactly?** First let me give the broad description
from 10,000ft up: Basically its a model for multinomial count
data (e.g., each sample contains the counts of $D$ "types of things"). Importantly, 
unlike the more common Poisson count models, the multinomial models a "competition to
be counted" (i.e., cases in which counting more of one type of thing means 
that I have less resources available to count other types of things). 

This may seem vague so let me give an example. Pretend there is a ball pit with 
red, green, and blue balls. Pretend that the ball pit is very large and I don't know 
the total number of balls in the ball pit, yet I want to say something about
the relative number of red, blue, and green balls in the pit. One way I may 
choose to measure the ball pit is by grabbing an armful of balls and counting 
the number of balls of each color (e.g., in one armful I may collect
5 red, 3 blue, and 6 green). My arms can only contain so many balls (in this example
about 14) and so if I were to have (randomly) gotten another green ball in my armful 
(making 7 total) I would likely not have been able to measure one of the red or blue balls;
hence the "competition to be counted". It turns out that this type of sampling
occurs all the time in many situations (Wikipedia has an example with [political 
polling](https://en.wikipedia.org/wiki/Multinomial_distribution#Example)). 
Perhaps one of the most notable examples of this type of count data occurs
with modern high-throughput sequencing studies such as 16S rRNA studies to 
profile microbial communities or bulk/single-cell RNA-seq studies to study 
expression profiles of cells. In all cases, transcripts are sequenced 
and the number of different types of transcripts are counted. The important part
is that sequencing only samples a small portion of the total genetic material 
available and leads to similar competition to be counted. 

## The *pibble* model
*Pibble* is one type of *stray* model. In particular its a *stray* model for 
multivariate linear regression. 

Let $Y$ denote an $D\times N$ matrix of counts. Let us denote the $j$-th 
column of $Y$ as $Y_j$. Thus each "sample" in the dataset is a measurement
of the relative amount of $D$ "types of things". Suppose we also 
have have covariate information in the form of a $Q\times N$ matrix $X$. 

The following is the pibble model including likelihood and priors:
$$
\begin{align}
Y_j & \sim \text{Multinomial}\left(\pi_j \right)  \\
\pi_j & = \phi^{-1}(\eta_j) \\
\eta_j &\sim N(\Lambda X_j, \Sigma) \\
\Lambda &\sim  MN_{(D-1) \times Q}(\Theta, \Sigma, \Gamma) \\
\Sigma &\sim W^{-1}(\Xi, \upsilon) 
\end{align}
$$
Here $MN_{(D-1) \times Q}$ denotes a [Matrix Normal distribution](https://en.wikipedia.org/wiki/Matrix_normal_distribution)
for a matrix $\Lambda$ of regression coefficients of dimension $(D-1)\times Q$. 
Essentially you can think of the Matrix normal as having two covariance matrices
one describing the covariation between the rows of $\Lambda$ ($\Sigma$) and another
describing the covariation of the columns of $\Lambda$ ($\Gamma$). 
and $W^{-1}$ refers to the [Inverse Wishart distribution](https://en.wikipedia.org/wiki/Inverse-Wishart_distribution) 
(which is a common distribution over covariance matrices).
The line $\pi_j = \phi^{-1}(\eta_j)$ represents a transformation between
the parameters $\pi_j$ which exist on a simplex (e.g., $\pi_j$ must sum to 1) and
the transformed parameters $\eta_j$ that exist in real space. In particular 
we define $\phi^{-1}$ to be the [inverse additive log ratio transform](http://www.sediment.uni-goettingen.de/staff/tolosana/extra/CoDaNutshell.pdf) (which conversely
implies that $\eta_j = ALR(\pi_j)$) also known as the identified softmax transform
(as it is more commonly known in the Machine Learning community). While 
I will say more on this later in this tutorial, one thing to know is that
I have the model implemented using the ALR transform as it is computationally
simple and fast; the results of the model can be viewed as if any number of 
transforms had been used (instead of the ALR) including the isometric log-ratio transform, or the 
centered log-ratio transform. 


Before moving on, I would like to give **a more intuitive description of *pibble***.
Essentially the main modeling component of *pibble* is the third equation above 
($\eta_j \sim N(\Lambda X_j, \Sigma)$) which is just a multivariate linear model. 
That is, $X$ are your covariates (which can be continuous, discrete, binary, etc...), 
and $\Sigma$ is the covariance matrix for the regression residuals.  


# Example analysis of microbiome data
This analysis is the same as that presented in the *stray* manuscript [@silverman2019]. 
I will reanalyze a previously published study comparing microbial composition in the terminal ileum of subjects with 
Crohn's Disease (CD) to healthy controls [@gevers2014]. To do this I will fit a pibble model using 
CD status, inflammation status and age as covariates (plus a constant intercept term). 

```{r message=FALSE, warning=FALSE}
library(MicrobeDS)
library(phyloseq)
library(dplyr)
library(stray)

set.seed(899)

data("RISK_CCFA")
# drop low abundant taxa and samples
dat <- RISK_CCFA %>% 
  subset_samples(disease_stat!="missing", 
                 immunosup!="missing") %>% 
  subset_samples(diagnosis %in% c("no", "CD")) %>% 
  subset_samples(steroids=="false") %>% 
  subset_samples(antibiotics=="false") %>% 
  subset_samples(biologics=="false") %>% 
  subset_samples(biopsy_location=="Terminal ileum") %>% 
  tax_glom("Family") %>% 
  prune_samples(sample_sums(.) >= 5000,.) %>%
  filter_taxa(function(x) sum(x > 3) > 0.10*length(x), TRUE)
```


Create Design Matrix and OTU Table
```{r}
sample_dat <- as.data.frame(as(sample_data(dat),"matrix")) %>% 
  mutate(age = as.numeric(as.character(age)),
         diagnosis = relevel(factor(diagnosis, ordered = FALSE), ref="no"), 
         disease_stat = relevel(factor(disease_stat, ordered = FALSE), ref="non-inflamed"))
X <- t(model.matrix(~diagnosis + disease_stat+age, data=sample_dat))
Y <- otu_table(dat)

# Investigate X and Y look like
X[,1:5]
Y[1:5,1:5]
```

Next specify priors. We are going to start by specifying 
a prior on the covariance between log-ratios $\Sigma$. I like to do this by 
thinking about a prior on the covariance between taxa on the log-scale (i.e.,
between the log of their absolute abundances not the log-ratios). I will refer
to this covariance on log-absolute abundances $\Omega$. For example, 
here I will build a prior that states that the mean of $\Omega$ is the identity matrix
$I_D$. From From @aitchison1986, 
we know that if we assume that the taxa have a covariance $\Omega$ in terms of log-absolute
abundance then their correlation in the $\text{ALR}_D$ is given by 
$$ \Sigma = G \Omega G^T $$
where $G$ is a $D-1 \times D$ matrix given by $G = [I_{D-1}; -1_{D-1}]$ (i.e.,
$G$ is the $\text{ALR}_D$ contrast matrix). Additionally, we know that
the Inverse Wishart mode is given by $\frac{\Xi}{\upsilon + D}$. Finally, 
note that $\upsilon$ essentially controls our uncertainty in $\Sigma$ about
this prior mean. Here I will take $\upsilon = D+3$. This then gives us 
$\Xi = (\upsilon - D) GIG^T$. We scale $\Xi$ by a factor of 1/2 to make $Tr(\Xi)=D-1$. 
```{r fig.height=5, fig.width=7}
upsilon <- ntaxa(dat)+3 
Omega <- diag(ntaxa(dat))
G <- cbind(diag(ntaxa(dat)-1), -1)
Xi <- (upsilon-ntaxa(dat))*G%*%Omega%*%t(G)
```

Finally I specify my priors for $\Theta$ (mean of $\Lambda$) and $\Gamma$ (covariance
between columns of $\Lambda$; i.e., covariance between the covariates). I will
center my prior for $\Lambda$ about zero, and assume that the covariates are independent. 
```{r fig.height=5, fig.width=7}
Theta <- matrix(0, ntaxa(dat)-1, nrow(X))
Gamma <- diag(nrow(X))
```

I strongly recommend users perform prior predictive checks to make sure their
priors make sense to them. *stray* makes this easy, all the main fitting functions 
(e.g., `pibble`) will automatically sample from the prior predictive 
distribution if `Y` is left as `NULL` (e.g., without data your posterior is
just your prior). 
```{r}
priors <- pibble(NULL, X, upsilon, Theta, Gamma, Xi)  
print(priors)
```

The main fitting functions in the *stray* package output special fit objects (e.g.,
`pibble` outputs an object of class `pibblefit`). These fit objects
are just lists with some extra metadata that allows special method dispatch. For 
example, if you call print on a `pibblefit` object you will get a nice 
summary of what is in the object. 


*Note:* Currently, the function `pibble` takes expects inputs and outputs in the "default" coordinate system;
this is simply the ALR coordinate system where the last category (`r ncategories(priors)` above) is taken as 
reference (this will be generalized in future versions). 
More specifically for a vector $x$ representing the proportions of 
categories $\{1, \dots, D\}$ we can write 
$$x^* = \left( \log \frac{x_1}{x_D}, \dots, \log \frac{x_{D-1}}{x_D}\right).$$
As mentioned above however, I have designed *stray* to work with many 
different coordinate systems including the ALR (with respect to any category), CLR, 
ILR, or proportions. To help transform things between these coordinate systems
I have written a series of transformation functions that transform any `pibblefit` 
object into a desired coordinate system. Importantly, `pibblefit` objects
keep track of what coordinate system they are currently in so as a user you only
need to specify the coordinate system that you want to change into. Keep in mind that
covariance matrices cannot be represented in proportions and so visualizations
or summaries based on covariance matrices will be suppressed when `pibblefit` objects
are in the proportions coordinate system. As an example, lets look at viewing
a summary of the prior for $\Lambda$ with respect to the CLR coordinate system^[These are 
very large objects with many posterior samples, so it can take a little time to compute. 
Faster implementations of summary may be included as a future update if need arises]. 

```{r}
priors <- to_clr(priors)  
summary(priors, pars="Lambda")  
```

By default the `summary` function returns a list (with possible elements `Lambda`, 
`Sigma`, and `Eta`) summarizing each posterior parameter based on quantiles and mean (e.g., 
p2.5 is the 0.025 percentile of the posterior distribution). As this type of 
table may be hard to take in due to how large it is, `pibblefit` objects also come with a default
plotting option for each of the parameters. Also the returned plot objects
are `ggplot` objects so normal `ggplot2` commands work on them. Before doing that
though we are going to use one of the `names` functions for `pibblefit` objects
to provide some more specific names for the covariates (helpful when we then plot). 

```{r fig.height=5, fig.width=7}
names_covariates(priors) <- rownames(X)
plot(priors, par="Lambda") + ggplot2::xlim(c(-10, 10))  
```

This looks fairly reasonable to me. So I am going to go ahead and 
fit the model with data. `stray` provides a helper method called `refit` 
that we will use to avoid passing prior parameters again. 

```{r}
priors$Y <- Y # remember pibblefit objects are just lists
posterior <- refit(priors, optim_method="adam")
```

Unlike the main *pibble* function, the `refit` method can be called
on objects in any coordinate system and all transformations to and from the
default coordinate system are handled internally^[That said, due to the need to transform
back and forth from the default coordinate system, it is fastest to call refit on 
`pibblefit` objects in the default coordinate system bypassing these transforms.]. 
This is one nice thing about
using the `refit` method. That said, new objects added to the `pibblefit` object 
need to be added in the proper coordinates For example, if we wanted to 
replace our prior for $\Xi$ for an object in CLR coordinates, we would
had to transform our prior for `Xi` to CLR coordinates before adding it to the `priors` object. 

Now I are also going to add in the taxa names to make it easier to 
interpret the results. 
```{r}
tax <- tax_table(dat)[,c("Class", "Family")]
tax <- apply(tax, 1, paste, collapse="_")
names_categories(posterior) <- tax
```

Before doing anything else lets look at the posterior predictive distribution
to assess model fit. This can be accessed through the method `ppc`^[This can also 
be used to plot samples of the prior predictive distribution if Y is null in the 
object as in our `priors` object].

```{r fig.height=5, fig.width=7}
ppc(posterior) + ggplot2::coord_cartesian(ylim=c(0, 30000))
```

There are a few things to note about this plot. First, when zoomed out like this
it looks it is hard to make much of it. This is a fairly large dataset we are 
analyzing and its hard to view an uncertainty interval; in this case its plotting
the median and 95% confidence interval in grey and black and the observed counts in green.
*stray* also has a simpler function that summarizes the posterior predictive check. 

```{r}
ppc_summary(posterior)
```

Here we see that the model appears to be fitting well (at least based on the posterior
predictive check) and that only about 1.5% of observations fall outside of the 95% 
posterior predictive density (this is good). 

Some readers will look at the above `ppc` plots and think "looks like over-fitting". 
However, note that there are two ways of using `ppc`. One is to predict the counts
based on the samples of $\eta$ (Eta; as we did above); the other is to predict "from scratch"
that is to predict starting form the posterior samples of $\Lambda$ (Lambda) then 
sampling $\eta$ and only then sampling $Y$. This later functionality can be accessed
by also passing the parameters `from_scratch=TRUE` to the `ppc` function. Note:
these two posterior predictive checks have different meanings, one is not better 
than the other.
```{r}
ppc(posterior, from_scratch=TRUE) +ggplot2::coord_cartesian(ylim=c(0, 30000))
ppc_summary(posterior, from_scratch=TRUE)
```

Now we are going to finally look at the posterior distribution of our regression
parameters, but because there are so many we will focus on just those 
that have a 95% credible interval not including zero (i.e., those that 
the model is fairly certain are non-zero). We are also going to ignore the 
intercept term and just look at parameters associated with age and disease status. 

```{r fig.height=5, fig.width=7}
posterior_summary <- summary(posterior, pars="Lambda")$Lambda
focus <- posterior_summary[sign(posterior_summary$p2.5) == sign(posterior_summary$p97.5),]
focus <- unique(focus$coord)
plot(posterior, par="Lambda", focus.coord = focus, focus.cov = rownames(X)[2:4])
```

The first, and most obvious ting to notice is that the covariate `age` has pretty 
much no effect at all, whatever effect it may have is incredibly weak. So 
we are going to remove age from the plot and just look at those 
coordinates with non-zero effect for diagnosis CD

```{r fig.height=5, fig.width=7}
posterior_summary <- filter(posterior_summary, covariate=="diagnosisCD") 
focus <- posterior_summary[sign(posterior_summary$p2.5) == sign(posterior_summary$p97.5),]
focus <- unique(focus$coord)

tax_table(dat)[taxa_names(dat)[which(names_coords(posterior) %in% focus)]]
plot(posterior, par="Lambda", focus.coord = focus, focus.cov = rownames(X)[2])
```


# More Technical Details 
## A few notes on model inference and parameter collapsing

Along with some algorithmic speed-ups enabled by the C++ Eigen library *stray* uses conjugate priors for the regression component of the model allowing the last three lines of the model to be
collapsed into 1 line. After this the last three lines of the model can be re-expanded using fully conjugate sampling schemes
that do not require optimization or MCMC (only matrix operations). 

**Here are the details:** The collapsed model is given by 
$$
\begin{align}
Y_j & \sim \text{Multinomial}\left(\pi_j, n_j\right)  \\
\pi_j & = \phi^{-1}(\eta_j) \\
\eta_j &\sim T_{(D-1)\times N}(\upsilon, \Theta X, \Xi, I_N + X^T \Gamma X)
\end{align}
$$
where $A=(I_N + X^T \Gamma, X)^{-1}$ and $T_{(D-1)\times N}$ refers to the Matrix T-distribution the $(D-1)\times N$ matrix $\eta$ with log density given by 
$$\log T_{(D-1)\times N}(\eta | \upsilon, \Theta X, \Xi, A) \propto -\frac{\upsilon+N-D-2}{2}\log | I_{D-1}+\Xi^{-1}(\eta-\Theta X)A(\eta-\Theta X)^T |.$$
Rather than using MCMC to sample $\eta$ stray uses MAP estimation (using a custom C++ Eigen based implementation of the ADAM optimizer and closed form solutions for gradient and hessian of the collapsed model)^[Which we found preformed substantially better than L-BFGS, which we also tried.]. Additionally, *stray* allows quantification of uncertainty in MAP estimates using a Laplace approximation. We found that in practice this MAP based Laplace approximation produced 
comparable results to a full MCMC sampler but with tremendous improvements in compute time. 

Once samples of $\eta$ are produced using the Laplace approximation closed form 
solutions for the conditional density of $\Lambda$ and $\Sigma$ given $\eta$ are 
used to "uncollapse" the collapsed model and produce posterior samples from the target 
model. This uncollapsing is fast and given by the following matrix equations:

$$
\begin{align}
\upsilon_N &= \upsilon+N \\
\Gamma_N &= (XX^T+\Gamma^{-1})^{-1} \\
\Theta_N &= (\eta X^T+\Theta\Gamma^{-1})\Gamma_N \\
\Xi_N &= \Xi + (\eta - \Theta_N X)(\eta - \Theta_N X)^T + (\Theta_N - \Theta)\Gamma(\Theta_N- \Theta)^T \\
p(\Sigma | \eta, X) &= W^{-1}(\Xi_N, \upsilon_N)\\
p(\Lambda | \Sigma, \eta, X) &= MN_{(D-1)\times Q}(\Lambda_N, \Sigma, \Gamma_N).
\end{align}
$$
If Laplace approximation is too slow, unstable (see below) or simply not needed, 
the default behavior of *pibble* is to preform the above matrix calculations and
produce a single point estimate of $\Sigma$ and $\Lambda$ based on the posterior
means of $p(\Sigma | \eta, X)$ and $(\Lambda | \Sigma, \eta, X)$. 


# References
