---
title: "Non-linear models with stray::basset"
author: "Justin Silverman"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{intro_stray_basset}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

If you have not already done so, I would read through the *pibble* vignette before this one.  

# stray::basset, the model
Stray can be used for non-linear modeling. Here we are going to go through one 
such model which is based on multivariate Gaussian processes. As Gaussian
processes are a type of lazy learner, I have named this model after a lazy dog: 
a Basset Hound. Hence this model is called *basset*. The *basset* model
can be written as 
$$
\begin{align}
Y_j & \sim \text{Multinomial}\left(\pi_j \right)  \\
\pi_j & = \phi^{-1}(\eta_j) \\
\eta &\sim N(\Lambda[X], \Sigma, I_N) \\
\Lambda[X] &\sim  \textsf{GP}(\Theta[X], \Sigma, \Gamma[X]) \\
\Sigma &\sim W^{-1}(\Xi, \upsilon) 
\end{align}
$$
where we now label quantities interpreted as function with square brackets. 
In particular, we use the notation that $\Lambda[X]$ is a function that maps am $Q \times N$ 
matrix of covariates $X$ to a matrix $\eta$. $\Lambda$ can be interpreted
as a smooth function that relates covariates $X$ to composition $\eta$ in a potentially
non-linear manner. $\Theta[X]$ is the mean function evaluated at $X$, 
$\Gamma[X]$ is a covariance matrix (or gram matrix) formed by evaluating
a positive semi-definite kernel function $K$ at each set of points $i$ and $j$, 
*i.e.*, $\Gamma_{ij} = K(X_i, X_j)$. Many different possible kernel 
functions makes this a very flexible and powerful model. 

# Non-linear time-series modeling
To demonstrate *basset* I am going to reanalyze the mallard^[I have a long
history of coming up with ridiculous names for statistical models and software.] dataset from 
@silverman2018 which is provided as a phyloseq object in *stray*. This dataset
features 4 artificial gut vessels sampled both daily and hourly for 1 month with 
many technical replicates. 

To accord with prior analyses (and to make it easy to visualize results) I am 
going to just analyze those bacterial families that are frequently observed. 
Note: the full dataset (at the sequence variant level without preprocessing)
is available if you run `data(mallard)`. For simplicity we will just look at 
samples from vessel 1^[In the future I may make a vignette discussing
how we can model all the vessels jointly using *basset*. Note: it's
just a different kernel function that also relates the covariance between
samples as a function of which vessel they were taken from.]. 

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(stray)
library(phyloseq)
library(dplyr)
library(driver)
library(ggplot2)
data(mallard_family)

# Just take vessel 1
mallard_family <- prune_samples(sample_data(mallard_family)$Vessel==1, mallard_family)

# Just take hourly samples
mallard_family <- prune_samples((sample_data(mallard_family)$time > "2015-11-20 15:00:00 UTC") &
                      (sample_data(mallard_family)$time < "2015-11-25 16:00:00 UTC"), mallard_family)

# Order samples - to make plotting easy later
o <- order(sample_data(mallard_family)$time)
otu_table(mallard_family) <- otu_table(mallard_family)[o,]
sample_data(mallard_family) <- sample_data(mallard_family)[o,]

# Extract Data / dimensions from Phyloseq object
Y <- t(as(otu_table(mallard_family), "matrix"))
rownames(Y) <- taxa_names(mallard_family)
D <- ntaxa(mallard_family)
N <- nrow(sample_data(mallard_family))

# X in hours
X <- as.numeric(sample_data(mallard_family)$time)
X <- t((X-min(X)) / 3600)
```

The function *basset* is a wrapper around some *stray* internals that make fitting
the above model easy. *basset* is patterned after the function *pibble* but
requires that `Theta` and `Gamma` be given as matrix functions rather than as
matrices. That is `Theta` must be a function that given `X` outputs a $D-1 \times N$ matrix 
(just like you would give to *pibble*) and `Gamma` must be a function that given `X`
outputs a $N \times N$ covariance matrix (e.g., covariance over the samples). 
There are a few kernel functions that meet the requirements for `Gamma` in *stray* already. 
Here we will use a Squared Exponential (SE) kernel for `Gamma` and will set `Theta`
to be the zero function. The SE kernel will give us non-linear smoothing of the
observed time-series. 

*Important Note:* Currently, `Theta` must output a matrix represented in the default coordinate 
system ($ALR_D$). This may be generalized in the future. 

Here we will just specify the Kernel parameters manually, more generally these
parameters can be chosen by cross-validation. 

```{r}
# Specify Priors
Gamma <- function(X) SE(X, sigma=5, rho=10) # Create partial function 
Theta <- function(X) matrix(0, D-1, ncol(X))
upsilon <- D-1+3
Xi <- matrix(.4, D-1, D-1)
diag(Xi) <- 1

# Now fit the model
fit <- stray::basset(Y, X, upsilon, Theta, Gamma, Xi)
```

It turns out that *basset* is really just creating a *pibblefit* object that 
is a little special because it represents a posterior over non-linear functions. 
The benefit of this is that many of the methods available for *pibblefit* objects
work for *bassetfit* objects. For example, we can use the same transformation methods:
```{r fig.height=4, fig.width=6}
fit.clr <- to_clr(fit)

# Plot Sigma in CLR
plot(fit.clr, par="Sigma", focus.coord=c("clr_seq_6", "clr_seq_5", "clr_seq_2"))
```

Really *basset* shows its power when you use it to smooth. 
```{r}
# predict not just missing days but also forecast into future
X_predict <- t(1:(max(X)))
predicted <- predict(fit.clr, X_predict, jitter=1) 
```

Now I am going to create a visual that shows the observed data in CLR coordinates 
(to do that I will need to add a pseudo-count) along with the smoothed estimates. 
```{r fig.height=5, fig.width=7}
family_names <- as(tax_table(mallard_family)[,"Family"], "vector")
Y_clr_tidy <- clr_array(Y+0.65, parts = 1) %>% 
  gather_array(mean, coord, sample) %>% 
  mutate(time = X[1,sample], 
         coord = paste0("CLR(", family_names[coord],")"))

predicted_tidy <- gather_array(predicted, val, coord, sample, iter) %>% 
  mutate(time = X_predict[1,sample]) %>% 
  filter(!is.na(val)) %>% 
  group_by(time, coord) %>% 
  summarise_posterior(val, na.rm=TRUE) %>% 
  ungroup() %>% 
  mutate(coord = paste0("CLR(", family_names[coord],")"))

ggplot(predicted_tidy, aes(x = time, y=mean)) +
  geom_ribbon(aes(ymin=p2.5, ymax=p97.5), fill="darkgrey", alpha=0.5) +
  geom_ribbon(aes(ymin=p25, ymax=p75), fill="darkgrey", alpha=0.9)+
  geom_line(color="blue") +
  geom_point(data = Y_clr_tidy, alpha=0.5) +
  facet_wrap(~coord, scales="free_y") +
  theme_minimal()+
  theme(axis.title.y = element_blank(), 
        axis.title.x = element_blank(), 
        axis.text.x = element_text(angle=45))

```


# References
